---
title: "FDA Baseball"
output: html_document
date: "2025-02-25"
---

# Driveline FDA analysis

Read in data and plot some raw data - here i've chosen shoulder angle z.

```{r}
# install.packages("fda") # install the fda package
# install.packages("refund") # install the refund package

library(fda)
library(refund)
library(tidyverse)
library(here)
library(magrittr)
library(viridis)

here::i_am("baseball_biomechanics/Analysis_Files/FDA Baseball.Rmd")

joint_angles <- read_csv(here("baseball_biomechanics/Data/joint_angles.csv"))
poi_metrics <- read_csv(here("baseball_biomechanics/Data/poi_metrics.csv"))
#landmarks <- read_csv(here("baseball_biomechanics/Data/landmarks.csv"))

pitches <- unique(joint_angles$session_pitch)

#graph of raw pitches (first 20)
joint_angles %>% 
        filter(session_pitch %in% pitches[1:20]) %>%
        #filter(session_pitch == "1031_2") %>%
        ggplot(aes(x = time, y = shoulder_angle_z, colour = session_pitch)) +
        geom_line(alpha = 0.6) +
        theme(legend.position = "none")

# extract the row when ball was released, and foot plant for each pitch, used in graph below
shoulder_at_br <- joint_angles %>% 
                  group_by(session_pitch) %>%
                  mutate(sh_rot_br = case_when(time == BR_time ~ "ball_release", 
                                               TRUE ~ NA)) %>%
                  filter(sh_rot_br == "ball_release")

shoulder_at_fp10 <- joint_angles %>% 
                  group_by(session_pitch) %>%
                  mutate(sh_rot_fp_10 = case_when(time == fp_10_time ~ "fp_10", 
                                               TRUE ~ NA)) %>%
                  filter(sh_rot_fp_10 == "fp_10")

shoulder_at_fp100 <- joint_angles %>% 
                  group_by(session_pitch) %>%
                  mutate(sh_rot_fp_100 = case_when(time == fp_100_time ~ "fp_100", 
                                               TRUE ~ NA)) %>%
                  filter(sh_rot_fp_100 == "fp_100")

# graph showing fp_10, fp_100, and ball release on each pitch. only first 5 pitches selected
joint_angles %>% 
        filter(session_pitch %in% pitches[c(1:5)]) %>%
        #filter(session_pitch == "1031_2") %>%
        ggplot(aes(x = time, y = shoulder_angle_z, colour = session_pitch)) +
        geom_line(alpha = 0.6) +
        geom_point(data = filter(shoulder_at_br, session_pitch %in% pitches[1:5]), aes(x = BR_time, y = shoulder_angle_z), colour = "black", alpha = 0.6) +
        geom_point(data = filter(shoulder_at_fp10, session_pitch %in% pitches[1:5]), aes(x = fp_10_time, y = shoulder_angle_z), colour = "red", alpha = 0.6) +
        geom_point(data = filter(shoulder_at_fp100, session_pitch %in% pitches[1:5]), aes(x = fp_100_time, y = shoulder_angle_z), colour = "blue", alpha = 0.6) +
        theme(legend.position = "none")

joint_angles %<>% select(session_pitch, time, shoulder_angle_z, fp_10_time, fp_100_time, MER_time, BR_time, MIR_time)

```

### Align pitches

## I commented this out as I will leave it to the registration step to align the pitches ## 

As the pitches are not aligned in time, transform the data so that max shoulder angle z occurs at time 0. This aligns the data.

```{r}
# 1. Find the time of the maximum joint angle for each pitch
# transformed_data <- joint_angles %>%
#   select(session_pitch, time, shoulder_angle_z) %>%
#   group_by(session_pitch) %>%
#   mutate(max_angle_time = time[which.max(shoulder_angle_z)],  # Find the time where the max angle occurs
#          time_aligned = time - max_angle_time) %>%        # Align all times so max is at time 0
#   ungroup()
# 
# transformed_data %<>% left_join(poi_metrics, by = "session_pitch")
# 
# transformed_data %>% 
#         filter(session_pitch %in% pitches[1:20]) %>%
#         ggplot(aes(x = time_aligned, y = shoulder_angle_z, colour = pitch_speed_mph, group = session_pitch)) +
#         geom_line(alpha = 0.5) +
#         scale_color_viridis()
# 
# # remove data early in the pitch 
# transformed_data %<>% filter(time_aligned > -0.25,
#                              time_aligned < 0.11)
# 
# transformed_data %>% 
#         filter(session_pitch %in% pitches[1:20]) %>%
#         ggplot(aes(x = time_aligned, y = shoulder_angle_z, colour = pitch_speed_mph, group = session_pitch)) +
#         geom_line(alpha = 0.5) +
#         scale_color_viridis()
```

### Convert data into FDA compatible matrix

A lot of copy/paste of Ed's workshop

```{r}

# # remove data early and late in the pitch based on ball release time
# this also does a lot of the work to align the pitches. 
joint_angles_trimmed <- joint_angles %>% group_by(session_pitch) %>%
  mutate(time_from_br = time - BR_time,
         time_from_fp = time - fp_10_time) %>%
  filter(time_from_fp > -0.5, time_from_br < 0.2) %>%
  filter(!session_pitch %in% c("2918_4", "2918_3")) #this pitch has a weird look about it. 

# filter out any pitches that have NA in fp_10_time (there were a few)
joint_angles_trimmed <- joint_angles_trimmed %>% group_by(session_pitch) %>%
  filter(!is.na(fp_10_time))

# check trimmed data
# replicating 
joint_angles_trimmed %>% 
        filter(session_pitch %in% pitches[1:20]) %>%
        #filter(session_pitch == "1031_2") %>%
        ggplot(aes(x = time, y = shoulder_angle_z, colour = session_pitch)) + 
        geom_line(alpha = 0.6) +
        geom_point(data = filter(shoulder_at_br, session_pitch %in% pitches[1:20]), aes(x = BR_time, y = shoulder_angle_z), colour = "black", alpha = 0.6) +
        theme(legend.position = "none")

#convert to a matrix
fda_matrix <- joint_angles_trimmed %>% 
  mutate(time = 1:length(session_pitch)) %>%
  #group_by(session_pitch) %>%
  #mutate(std_time = 1:length(session_pitch)) %>%
  #ungroup() %>%
  select(session_pitch, time, shoulder_angle_z) %>%
  pivot_wider(names_from = session_pitch, values_from = shoulder_angle_z)

fda_matrix <- as.matrix(fda_matrix)
```


```{r}

t_grid <- as.numeric(fda_matrix[,1])
row.names(fda_matrix) <- t_grid #change row name sto t_grid (may be an unnessessary step?)
fda_matrix <- fda_matrix[1:423,2:402] # starts at 2 to exclude the time column
t_grid <- as.numeric(1:423) 


matplot(x = t_grid, 
        y = fda_matrix,
        type = "b", # both lines and points
        pch = 20, # dots for points
        xlab = "Normalised time",
        ylab = "Shoulder angle (deg)")

# bspl_101 <- create.bspline.basis(rangeval = range(t_grid), nbasis = 100, norder = 4)
# ols_fdSmooth <- smooth.basis(argvals = t_grid, y = fda_matrix, fdParobj = bspl_101)
# ols_fd <- ols_fdSmooth$fd # or ols_fdSmooth[["fd]]
# plot(ols_fd[1:3,],
#      xlab = "Normalised time (prop. of gait cycle)",
#      ylab = "Hip angle (deg)")
# matpoints(x = t_grid, y = fda_matrix[, 1:3], pch = 20) # overlay
```

```{r}
# plot(t_grid, fda_matrix[,1],  # plot raw data (1st curve)
#      pch = 20,
#     xlab = "Normalised time (prop. of gait cycle)",
#     ylab = "Hip angle (deg)")
     
# K_seq <- seq(4, 20, by = 2)

# for(Ki in seq_along(K_seq)) {
#   bspl_K <- create.bspline.basis(rangeval = range(t_grid), nbasis = K_seq[Ki], norder = 4)
#   ols_fdSmooth_K <- smooth.basis(argvals = t_grid, y = fda_matrix[,1], fdParobj = bspl_K)
#   lines(ols_fdSmooth_K$fd, col = Ki)
# }
# legend("top", legend = paste0("K = ", K_seq), col = seq_along(K_seq), lty = 1)
```

### P-OLS

```{r}
# create a new smooth.basis object with penalised OLS
bspl_100 <- create.bspline.basis(rangeval = range(t_grid), nbasis = 100, norder = 4) # change nbasis values
fdParobj_pols <-fdPar(fdobj = bspl_100, Lfdobj = int2Lfd(m = 2), lambda = 1)
pls_fdSmooth <- smooth.basis(argvals = t_grid, y = fda_matrix, fdParobj = fdParobj_pols)


# Get best lamba value for the above function
# create vcector to store the GCV values:
log10_lambda_range <- seq(-10, 6, by = 2)
lambda_range <- 10^log10_lambda_range
n_lambda <- length(lambda_range) # how many values we're trialling
gcv_vec <- vector(mode = "numeric", length = n_lambda)

for(lam in seq_len(n_lambda)) {
  # loop through different lambdas
  fdParobj_lam <- fdPar(fdobj = bspl_100, 
                        Lfdobj = int2Lfd(m = 2),
                        lambda = lambda_range[lam]) # create fdPar with chosen lambda
  # do p-ols with chosen lambda:
  pls_fdSmooth_lam <- smooth.basis(argvals = t_grid, y = fda_matrix, fdParobj = fdParobj_lam)
  # store resulting GCV:
  gcv_vec[lam] <- mean(pls_fdSmooth_lam$gcv)
}

plot(log10_lambda_range, 
     y = gcv_vec, 
     type = "b", 
     xlab = expression(log[10](lambda)), 
     ylab = "GCV")
best_lambda_index <- which.min(gcv_vec)
abline(v = log10_lambda_range[best_lambda_index])
```

Implement the lamba value found above using generalised CV

```{r}
fdParobj_final <- fdPar(fdobj = bspl_100, 
                        Lfdobj = int2Lfd(m = 2),
                        lambda = lambda_range[best_lambda_index]) # create fdPar with chosen 
pls_fdSmooth_final <- smooth.basis(argvals = t_grid, y = fda_matrix, fdParobj = fdParobj_final)
final_fd <- pls_fdSmooth_final$fd
```


```{r}
# define coarse grid
t_grid_coarse <- seq(min(t_grid), max(t_grid), length.out = 10)
# evaluate on that grid
eval.fd(evalarg = t_grid_coarse, fdobj = final_fd)

mean_fd <- mean.fd(x = final_fd)
final_coef <- final_fd$coef # extract coefficients
mean_coef <- apply(final_coef, 1, mean) # average them
mean_fd_02 <- fd(coef = mean_coef, basisobj = final_fd$basis) # create mean fd object manually

plot(final_fd)
lines(mean_fd, lwd = 3)
lines(mean_fd_02, lwd = 3)

boxplot.fd(x = final_fd)
```

## Registration

Need to get the landmark registration data. IE the foot plant time and ball release time. 

```{r}
#get the angle and time where fp occurs
lm_fp <- joint_angles_trimmed %>% group_by(session_pitch) %>%
  mutate(time_std = 1:length(time)) %>%
  filter(time == fp_10_time)
#lm_fp <- lm_fp[1:99,]

# get the angle and time when br occurs
lm_br <- joint_angles_trimmed %>% group_by(session_pitch) %>%
  mutate(time_std = 1:length(time)) %>%
  filter(time == BR_time)
#lm_br <- lm_br[1:99,]

# Replicating the plot in the registration workshop practical (picking out the landmarks)
t_min <- as.numeric(unlist(lm_fp[, 10]))
y_min_value <- as.numeric(unlist(lm_fp[, 3]))
t_max <- as.numeric(unlist(lm_br[, 10]))
y_max_value <- as.numeric(unlist(lm_br[, 3]))

inds_obs_plot <- c(1, 11, 21) # indices of three observations to plot
plot(final_fd[inds_obs_plot], xlab = "t", ylab = "force (bw)")
points(t_min[inds_obs_plot], y_min_value[inds_obs_plot], col = c(1:3), cex = 1.5, pch = 20)
points(t_max[inds_obs_plot], y_max_value[inds_obs_plot], col = c(1:3), cex = 1.5, pch = 20)


#Set the foot plant and ball release data into a matrix
ximarks_mat <- matrix(data = NA, nrow = 402, ncol = 2) # set up empty matrix
ximarks_mat[, 1] <- lm_fp$time_std
ximarks_mat[, 2] <- lm_br$time_std




# im not sure what this is doing. From the workshop - "A vector of length NL of times of landmarks for target curve – here as targets we’ll just use the mean average of each landmark timing"
x0marks_vec <- c(mean(lm_fp$time_std), mean(lm_br$time_std))


bspline_basis_wfd <- create.bspline.basis(rangeval = c(1, 423),
                                          breaks = c(1, mean(lm_fp$time_std), 
                                                     mean(lm_br$time_std), 423),
                                          norder = 4)

# Wfdpar (and Wfd0par): a functional parameter object defining the (inverse) warping functions. For this we first set up a B-spline basis with interior knots at the mean values of the two landmarks using the breaks argument.

WfdLM <- fd(coef = matrix(0, bspline_basis_wfd$nbasis, 1), 
            basisobj =  bspline_basis_wfd) # are these initial guesses

WfdParLM <- fdPar(fdobj = WfdLM, Lfdobj = int2Lfd(2), lambda = 1e-6)

# New landmark reg function
source(here("baseball_biomechanics/landmark-reg-updated.R"))

fast_reg <- landmarkreg_u(fdobj = final_fd,
                    ximarks = ximarks_mat,
                    x0marks = x0marks_vec,
                    WfdPar = WfdParLM,
                    monwrd = 0)

# landmark_reg_obj <- landmarkreg(
#   unregfd = final_fd, 
#   ximarks = ximarks_mat,
#   x0marks = x0marks_vec,
#   WfdPar = WfdParLM,
#   WfdPar0 = WfdParLM,
#   x0lim = c(0, 423),
#   ylambda = 1e-8)

```

#### Examining the registered curves

```{r}
# extract
reg_fd <- fast_reg$regfd
warp_fd <- fast_reg$warpfd

par(mfrow = c(1, 2))
plot(final_fd)

title("Unregistered curves")

plot(reg_fd)

abline(v = x0marks_vec, col = "grey")
title("Registered curves")

# replicating the previous 3 curve plot 

par(mfrow = c(1, 2))
plot(final_fd[inds_obs_plot, ])
points(t_min[inds_obs_plot], y_min_value[inds_obs_plot], col = c(1:3), cex = 1.5, pch = 20)
points(t_max[inds_obs_plot], y_max_value[inds_obs_plot], col = c(1:3), cex = 1.5, pch = 20)
title("Unregistered curves")

plot(reg_fd[inds_obs_plot,])

abline(v = x0marks_vec, col = "grey")
points(rep(mean(t_min), 3), y_min_value[inds_obs_plot], col = c(1:3), cex = 1.5, pch = 20)
points(rep(mean(t_max), 3), y_max_value[inds_obs_plot], col = c(1:3), cex = 1.5, pch = 20)
title("Registered curves")



```
#### Warping function 

```{r}
plot(warp_fd)
plot(warp_fd[inds_obs_plot,])
```

#### Analysis of phase variation

Note I use handedness (in place of sex from the practical), there are 27 L and 72 R in the sample. Ball release graph looks strange. this is because earlier I standardised the curves based on BR (ie curve starts and ends 0.4 sec before BR and 0.2 second after) so there is no variation in this. 

```{r}
par(mfrow = c(1, 2))
boxplot(t_min ~ poi_metrics$p_throws[1:99],
        col = c("red4", "cornflowerblue"),
        xlab = "handedness", 
        ylab = "landmark time: foot plant")
boxplot(t_max ~ poi_metrics$p_throws[1:99],
        col = c("red4", "cornflowerblue"),
        xlab = "sex", 
        ylab = "landmark time: ball release")
```


```{r}
plot(warp_fd[1,])
lines(warp_fd, col = ifelse(poi_metrics$p_throws[1:99] == "L", "red4", "cornflowerblue"), lty = 1)



```

## FPCA

```{r}
cov_grf_bifd <- var.fd(fdobj1 = warp_fd)
cov_grf_eval <- eval.bifd(sevalarg = t_grid, tevalarg = t_grid, bifd = cov_grf_bifd)
filled.contour(x = 1:178, y = 1:178, z = cov_grf_eval, xlab = "s", ylab = "t")
```

The graph shows pointwise variance, with the highest variance between 0 and 50, which makes sense as this is the phase before foot plant and mechanics of pitching can vary a lot at this stage. If we run the warp_fd through the graph we see the highest variance at 50-100. This is the area with the most warping. 

### Applying FPCA

```{r}
pitching_pca.fd <- pca.fd(fdobj = reg_fd, nharm = 6)
pitching_funs_fd <- pitching_pca.fd$harmonics

#The FPC scores are returned in an N x 6 matrix
pitching_fpc_scores <- pitching_pca.fd$scores

#The eigenvalues, which represent the amount of variance explained by the FPCs, are contained in a vector called values
pitching_eigenvalues <- pitching_pca.fd$values
# actually shows all possible values, we just want first 6
pitching_eigenvalues <- pitching_eigenvalues[1:6]

#The normalised eigenvalues that represent the proportion of variance explained are contained in the vector called varprop
pitching_varprop <- pitching_pca.fd$varprop
```

### Describing the FPCA

```{r}
info_df <- data.frame(`FPC Number` = 1:6,
           `Eigenvalue` = round(pitching_eigenvalues, 2),
           `Proportion of Variance` = round(pitching_varprop, 2),
           `Cumulative Proportion of Variance` = cumsum(round(pitching_varprop, 2)))

info_df
```

```{r}
par(mfrow = c(2, 3))
plot.pca.fd(x = pitching_pca.fd)
```

Relate these to actual pitch data showing pitches with highest and lowest FPC scores

```{r}
pitching_fpc1_score <- pitching_fpc_scores[,1]
max_ind <- which.max(pitching_fpc1_score)
min_ind <- which.min(pitching_fpc1_score)

par(mfrow = c(1, 2))
plot.pca.fd(x = pitching_pca.fd, harm = 1)
plot(reg_fd[c(min_ind, max_ind)])
legend("bottom", 
       col = c(1,2), 
       lty = c(1, 2), 
       legend = c("Min FPC1 score", "Max FPC1 score"),
       "Lowest and Highest FPC1 Scorers")
```

### FPCs as basis functions

```{r}
mean_pitching_fd <- mean.fd(x = reg_fd)
subject_1_scores <- pitching_fpc_scores[1, ]
new_coefs <- as.vector(subject_1_scores %*% t(pitching_funs_fd$coefs))
scores_times_fpcs_fd <- fd(coef = new_coefs, basisobj = pitching_funs_fd$basis)
approximation_subject_1_fd <- mean_pitching_fd + scores_times_fpcs_fd

plot(approximation_subject_1_fd)
lines(reg_fd[1,], lty = 2)
legend("bottom", c("True", "FPC Approx"), lty = c(2, 1))
```

## Functional regression
### Scalar-on-function regression 

```{r}
# outcome variable
pitch_speed <- poi_metrics$pitch_speed_mph[1:99]

# take the FPCA scores and combine with outcome variable into a new dataframe

fpcr_df <- data.frame(pitching_fpc_scores, pitch_speed)
names(fpcr_df)[1:6] <- paste0("fpc_", 1:6)
names(fpcr_df)[7] <- "pitch_speed"

#create the model
fpcr_lm <- lm(formula = pitch_speed ~ fpc_1 + fpc_2 + fpc_3 + fpc_4 + fpc_5 + fpc_6, 
   data = fpcr_df)
summary(fpcr_lm)
```

There is a better way

```{r}
library(refund)

pitch_fd_eval <- t(eval.fd(evalarg = 1:99, fdobj = reg_fd))
pfr_fit <- pfr(pitch_speed ~ lf(X = pitch_fd_eval, bs = "bs", k = 35, argvals = 1:99))
summary(pfr_fit)

plot(pfr_fit, xlab = "t", ylab = expression(beta(t)))
abline(h =0, lty = 3)
```

This does not do a great job of predicting pitch speed. The r-squared is 0.06. Basically no relationship at all. 

An additional step we can take is to include other joint metrics (angles, velocities etc) to build a multiple linear model. This could improve the prediction accuracy if the metrics are related to pitch speed. I guessed that rate of change in shoulder rotation (current data) would relate to pitch speed, it does not. What other metrics may influence speed? 


